The effect of habitat choice on evolutionary rescue in subdivided populations

### 2020-07-10 ms/main.tex
DL modifs done on overleaf

The probability of adaptation $P_{\mathrm{adapt}}$ calculated with eq.~\ref{eq:source_sink} is compared to simulation results in Fig.~\ref{fig:source_sink}. In spite of our approximations, the fit to simulation results is still very good. 
\linelabel{R1-33}As $P_{\mathrm{adapt}}$ includes the probabilities of establishment $\varphi_{k}$, here again, the shapes of the curves as function of the emigration probability $m$ depend on the fecundity of the mutant in old-habitat patches, $\omega^\text{old}_m$ (Figs.~\ref{fig:source_sink_m-35}, \ref{fig:source_sink_m-45}). Likewise, the rankings of the different dispersal schemes are comparable to the ones observed for the establishment probability. \linelabel{R2-35}The small differences in the ranking of the dispersal schemes between panels~\ref{fig:vary_m_est_old-45} and \ref{fig:source_sink_m-45} can be explained by the mutational input. Larger stationary populations sizes of the wild type in new habitats, $\widehat{N}_w^{\text{new}}$ result in more mutations during the observed time frame. This population size is altered by the different dispersal schemes, most importantly it is largest for the wild type having a bias towards old habitats, i.e. the Old-Old and the Old-New dispersal schemes \florence{It does not make intuitive sense why $\widehat{N}_w^{\text{new}}$ is higher with WT biased towards old... maybe then figures with $\widehat{N}_w^{\text{old}}$ and $\widehat{N}_w^{\text{new}}$ would help}.

@@

We identified two types of effects of the different dispersal schemes. First, by changing population densities in both habitat types, the dispersal schemes alter the growth rate of the mutant in old-habitat patches $a_{\text{old}}$. This is the primary reason for the ranking of the dispersal schemes\footnote{\florence{really need a figure!}}. Secondly, the different dispersal schemes affect the number of mutations arising in either habitat type. This has a minor effect on the probability of evolutionary rescue for the explored parameter range but is relevant when studying the origin of the successful mutant lineage \chg{(see also Fig.~\label{Sfig:natal_habitat})}. As the genetic background may vary across patches, the origin of a successful mutation will also affect which mutations will hitchhike with it. \linelabel{R1-54}Similarly, in the case of polygenic rescue or under recombination \citep[e.g.][]{uecker_2015}, the origin of a mutant is likely to affect its success.

@@

@@

@@

The mean reproduction matrix $\mathcal{M}$ of a mutant gives the average number of offspring in a certain habitat, dependent on the habitat type in which the mutant resides (see also eq.~\eqref{eq:mean_repro} in the main text):
\begin{equation}\label{eq:repromatrix}
\mathcal{M} = \bordermatrix{ ~ & \text{old patch} & \text{new patch} \cr
	\text{old patch} & \left(1 - m \mu_m^{\text{new}} \right) (1+a_{\text{old}}) & m \mu_m^{\text{new}} (1+a_{\text{new}}) \cr
	\text{new patch} & m \mu_m^{\text{old}} (1+a_{\text{old}}) & \left(1 - m \mu_m^{\text{old}}\right) (1+a_{\text{new}})
}\ ,
\end{equation}
\begin{equation}\label{Seq:mean_repro}
\mathcal{M} = \bordermatrix{ ~ & \text{old patch} & \text{new patch} \cr
	\text{old patch} & \left(1-m_m^{\text{new}}\right) (1+a_{\text{old}}) & m_m^{\text{new}} (1+a_{\text{new}}) \cr
	\text{new patch} & m_m^{\text{old}} (1+a_{\text{old}}) & \left(1-m_m^{\text{old}}\right) (1+a_{\text{new}})
}\ ,
\end{equation}
where the rows denote the parent locations, and the columns the patch type of the offspring.

Our goal is to apply Theorem 5.6 from \citet{haccou_book} which states that for a slightly super-critical branching process, \chg{i.e. where the survival probability is slightly above zero}, the establishment probability can be expressed in terms of the largest eigenvalue $\rho$ and the corresponding left- and right-eigenvectors of the mean reproduction matrix \chg{$\mathcal{M}$}, denoted by $u$ and $v$, respectively. The eigenvectors should be normalized in the following way: $u_1+u_2 = 1$ and $\sum_{i=1}^2 u_i v_i = 1$. The establishment probabilities are then given by 
\begin{equation}\label{Seq:theory}
\varphi_i = \frac{2(\rho-1)}{B} v_i + O(\varepsilon), 
\end{equation}
with
\begin{equation} 
B = \sum_{i=1}^2 u_i \sum_{j=1}^2 v_j \mathcal{M}_{ij} + \rho(1-\rho) \sum_{j=1}^2 u_j v_j^2\, . 
\end{equation}

\subsection{Computing the largest eigenvalue}
We first approximate the largest eigenvalue of $\mathcal{M}$ denoted by $\rho$. It is given by (see \textit{Mathematica} notebook)
\begin{equation}
\begin{aligned}
\rho &= \frac{1}{2}\left(2 + a_{\text{old}} + a_{\text{new}} - m -  m_m^{\text{new}} a_{\text{old}} - m_m^{\text{old}} a_{\text{new}} \right.\\
&\quad \left. + \sqrt{4 (m -1) (1+a_{\text{old}})(1+a_{\text{new}}) + (2+a_{\text{old}} + a_{\text{new}} - m - m_m^{\text{new}} a_{\text{old}} - m_m^{\text{old}} a_{\text{new}}})^2 \right)
\end{aligned}
\end{equation}
%
In order to make analytical progress and to identify under which conditions the process is slightly super-critical, i.e. $\rho>1$, we rescale the parameters by a small parameter $\varepsilon$. We set $a_{\text{old}} = \varepsilon, a_{\text{new}} = \varepsilon \xi$ and $m = \varepsilon\mu$. Assuming that $\varepsilon$ is small enough, i.e. effectively a weak selection assumption in old-habitat patches, we can neglect higher orders of $\varepsilon$ and find
\begin{equation}\label{eq:eigenvalue}
\begin{aligned}
\rho &\approx 1+\frac{1}{2}\varepsilon\left(1+ \xi - \mu + \sqrt{(\xi -1 + \mu_m^{\text{new}})^2 + 2(1-\xi+\mu_m^{\text{new}})\mu_m^{\text{old}} + (\mu_m^{\text{old}})^2  }\right)\\
&= 1+\frac{1}{2}\varepsilon\left(1+ \xi - \mu + \sqrt{\frac{\gamma}{1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}}}  }\right),
\end{aligned}
\end{equation}
where $\gamma$ is the rescaled version of the constant $C$ in the main text (eq.~\eqref{eq:normalization}), i.e.
\begin{equation}
\gamma = (1-f_{\text{old}})(\xi-1+\mu)^2 +  e^{\pi_m} f_{\text{old}} (\xi-1-\mu)^2\, .
\end{equation}
%
For $\varepsilon\to 0$ we find that $\rho\to 1$ \chg{(eq.~\eqref{eq:eigenvalue}) which means that the branching is slightly super-critical if $\rho>1$ and real. A sufficient condition for this to be true is}
\begin{equation}
1+\xi-\mu > 0 \quad \Leftrightarrow \quad a_{\text{old}} + a_{\text{new}} - m > 0
\end{equation}
%
\chg{In case that the branching process is not super-critical the establishment probability in eq.~\eqref{Seq:theory} becomes negative and as such is not a probability anymore. Hence, we can simply reject negative solutions of the establishment probability and by that implicitly justify that our approximation is valid.}%\\

@@

\subsection{Computing the establishment probability}
For the solution of eq.~\eqref{Seq:theory} it remains to compute the normalized eigenvectors. Their precise form is of not much insight. We therefore omit stating them explicitly but refer to the \textit{Mathematica} notebook. Solving eq.~\eqref{Seq:theory} to the first order of $\varepsilon$ we find

\begin{equation}
\begin{aligned}
\varphi_{\text{old}} &= \varepsilon + \frac{\varepsilon\left(1-\xi\right)}{\sqrt{\frac{\gamma}{(1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}})}}}  + \frac{\varepsilon\left(\mu_m^{\text{old}} -\mu_m^{\text{new}} + 2 \mu_m^{\text{new}}\xi  \right)}{\sqrt{\frac{\gamma}{(1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}})}}}\, ,\\
\varphi_{\text{new}} &= \varepsilon\xi + \frac{\varepsilon\xi \left(\xi-1\right)}{\sqrt{\frac{\gamma}{(1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}})}}} + \frac{\varepsilon\left(\xi \mu_m^{\text{new}} -\xi \mu_m^{\text{old}} + 2 \mu_m^{\text{old}}  \right)}{\sqrt{\frac{\gamma}{(1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}})}}}\, .
\end{aligned}
\end{equation}
%
Transforming back to the original variables and replacing $\gamma$ by the constant $C$ from the main text (eq.~\eqref{eq:normalization})
\begin{equation}
C = (1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}}) \left((1-f_{\text{old}})(a_{\text{new}}-a_{\text{old}}+m)^2 +  e^{\pi_m} f_{\text{old}} (a_{\text{new}}-a_{\text{old}}-m)^2\right)\, ,
\end{equation}
we obtain

\begin{equation}
\begin{aligned}
\varphi_{\text{old}} &= a_{\text{old}} + \frac{(1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}})a_{\text{old}} \left(a_{\text{old}}-a_{\text{new}}\right)}{\sqrt{C}}  + \\
&\qquad \qquad \qquad m \frac{\left( e^{\pi_m} f_{\text{old}} a_{\text{old}} - (1-f_{\text{old}}) a_{\text{old}} + 2 (1-f_{\text{old}}) a_{\text{new}}  \right)}{\sqrt{C}}\, ,\\
\varphi_{\text{new}} &= a_{\text{new}} + \frac{(1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}})a_{\text{new}} \left(a_{\text{new}} - a_{\text{old}}\right)}{\sqrt{C}} \\
&\qquad \qquad \qquad + m \frac{\left(a_{\text{new}} (1-f_{\text{old}}) - a_{\text{new}}  e^{\pi_m} f_{\text{old}} + 2 a_{\text{old}}  e^{\pi_m} f_{\text{old}}  \right)}{\sqrt{C}}\, .
\end{aligned}
\end{equation}
%
Slightly re-ordering the terms, this gives the establishment probability of a single mutant individual, eq.~\eqref{eq:estab_approx} in the main text:
\begin{equation}\label{Seq:estab_approx}
\begin{aligned}
\varphi_{\text{old}} &\approx \qquad \quad \; \; \; \; a_{\text{old}} \qquad \quad \;  \; +  \quad a_{\text{old}} \frac{\left(1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}}\right)}{\sqrt{C}}(a_{\text{old}}-a_{\text{new}}) \\
& \qquad \qquad + \frac{m}{\sqrt{C}} \left(a_{\text{new}}(1-f_{\text{old}}) + a_{\text{old}} e^{\pi_m} f_{\text{old}} - (a_{\text{old}}-a_{\text{new}})(1-f_{\text{old}})\right) \, ,\\
\varphi_{\text{new}} &\approx \underbrace{ a_{\text{new}}}_{\text{(1) local growth parameter}} +  \quad \underbrace{ a_{\text{new}} \frac{\left(1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}}\right)}{\sqrt{C}}(a_{\text{new}}-a_{\text{old}})}_{\text{(2) effect of the heterogeneous environment}} \\
& \qquad \quad \underbrace{+\; \frac{m}{\sqrt{C}}\left( a_{\text{new}}(1-f_{\text{old}}) + a_{\text{old}}  e^{\pi_m} f_{\text{old}} - (a_{\text{new}}-a_{\text{old}})  e^{\pi_m} f_{\text{old}} \right)\, .}_{\text{(3) effect of dispersal: new patches $+$ old patches $-$ loss to the other patch type }}
\end{aligned}
\end{equation}

\chg{For $m=0$ we see that $\varphi_{\text{old}}=0$, i.e. terms (1) and (2) cancel out. For the establishment probability in the new habitat we recover Haldane's result for the establishment probability of a slightly advantageous mutant: $\varphi_{\text{new}}=2a_{\text{new}}$ \citep{haldane_1927}.}

\subsection{Disentangling the contributions to the establishment probability \label{sec:SI_contributions}}
\chg{We now proceed to explain} the three regions of the establishment probability from Fig.~\ref{fig:vary_m_est}(a) in the main text. These were defined by: (i) an initial increase of the establishment probability at low dispersal rates $m$; (ii) a local maximum with a subsequent decrease of the establishment probability; (iii) an increase of the establishment probability for high dispersal rates.

\begin{figure}[t!]
	\centering
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=\linewidth]{fig2a.pdf}
		\caption{$\omega^\text{old}_m=1.35$ (large fecundity difference)}
	\end{subfigure}%
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=\linewidth]{fig2c.pdf}
		\caption{$\omega^\text{old}_m=1.45$ (small fecundity difference)}
	\end{subfigure}
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=\linewidth]{figS1c.pdf}
		\caption{$\omega^\text{old}_m=1.35$ (large fecundity difference)}
	\end{subfigure}%
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=\linewidth]{figS1d.pdf}
		\caption{$\omega^\text{old}_m=1.45$ (small fecundity difference)}
	\end{subfigure}
	\caption{\textbf{Contribution of the different terms in eq.~\eqref{Seq:estab_approx} to the establishment probability $\varphi_{\text{old}}$.} Subfigures (a,b) are the same as Figs.~\ref{fig:vary_m_est}(a,c) in the main text. They show the establishment probability for a single mutant individual arising in an old-habitat patch for varying emigration probabilities $m$. In subfigures (c,d) we plot the terms from eq.~\eqref{Seq:estab_approx} separately ($\pi_w=0.5,\pi_m=-0.5$). Term 1, the mutant growth rate in old habitats (solid), increases with increasing dispersal rates as a consequence of relaxed competition. Term 2, the environmental effect (dashed), captures the differences between the growth rates in the habitats. The larger the difference, the larger its contribution to the overall establishment probability. Term 3, the effect of dispersal (dotted), (largely) increases with increasing dispersal rates $m$. The sum of the three terms is plotted as a dash-dotted line.}
	\label{Sfig:contribution}
\end{figure}

For clarity, we re-plot Figs.~\ref{fig:vary_m_est}(a,c) in Fig.~\ref{Sfig:contribution}(a,b), respectively. We try to explain the ongoing processes for each of the regions through the approximations of the establishment probabilities in eq.~\eqref{Seq:estab_approx}, see also Fig.~\ref{Sfig:contribution}(c,d). Note that these explanations are only valid for the establishment probability of a mutant initially in an old-habitat patch, $\varphi_{\text{old}}$. For the intuition behind the shape of $\varphi_{\text{new}}$ we refer to the corresponding section in the main text.

Region (i) is explained by the positive effect of dispersal. Mutants disperse from old- to new-habitat patches where they \chg{have a higher growth rate}. This effect is mediated through the third term of the establishment probability in eq.~\eqref{Seq:estab_approx}. \chg{Note that the first two terms of the approximation cancel out for small emigration probabilities $m$.}
While the third term increases with increasing emigration rate $m$, the second term in eq.~\eqref{Seq:estab_approx} decreases, cf. Fig.~\ref{Sfig:contribution}. 
In the formula this is mediated through \chg{the increase of the local growth rate in the old-habitat, $a_{\text{old}}$, i.e.\ it becomes less negative. Then both factors of the second term, $a_{\text{old}}$ and the difference $(a_{\text{old}}-a_{\text{new}})$, increase (which in turn decreases term two). The }\chg{intuitive} reason is that due to larger emigration probabilities $m$, more individuals leave old habitats before the reproductive event. This relaxes competition in these patches and therefore increases the local growth rate of mutants in old-habitat patches.   
%
%The intuition is that habitat differences become less important under stronger population mixing -- recall that the dispersal rate $m$ appears in the constant $C$ in the denominator. This alone would explain the decreasing curve of the second term. Additionally though, the local growth rate in old-habitat patches, $a_{\text{old}}$, is linked to the population demography which changes for varying $m$. With larger dispersal rates, more individuals leave the densely populated old habitats. This results in alleviated competition pressure for the remaining individuals, thus increasing the local growth rate $a_{\text{old}}$. This decreases the difference of growth rates $a_{\text{old}}-a_{\text{new}}$. Therefore, the local maximum can be explained by the decreasing environmental influence and the increasing dispersal effect, the second and the third term in eq.~\eqref{eq:estab_approx}, respectively. Hence, region (ii), beginning with the local maximum, is dominated by the decreasing effect of the environment on the establishment probability. 
Finally, in region (iii) dispersal is so large that the population homogenizes. This results in even less competitive pressure in old-habitat patches. Eventually, this yields a positive growth rate $a_{\text{old}}$ (first term in eq.~\eqref{Seq:estab_approx}). Therefore, this region is driven by the local growth rate in old habitats.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Native habitat type %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\renewcommand{\theequation}{C\arabic{equation}}
\setcounter{equation}{0}  % reset counter

@@

@@

\section{Probability of establishment for large frequencies of old-habitat patches}
\chg{We plot the probability of establishment for a large frequency of old-habitat patches, $f_{\text{old}}$. As visible in Fig.~\ref{Sfig:vary_f} below, for high frequencies of old-habitat patches the probability of adaptation becomes very small, if not zero, for large emigration probabilities $m$. These high frequency of old-habitat patches are the patch configurations which mutants that were present before the first patch degrades, i.e. standing genetic variation mutants, experience. Therefore, it is very unlikely that the mutant, which will eventually rescue the population, was already present before the first degradation event.} This supports the explanation that homogenizing the population (large $m$) reduces the impact of standing genetic variation on the probability of evolutionary rescue, see Fig.~\ref{fig:sgv}(b) in the main text where $\pi_w=\pi_m=0$ was plotted.

\begin{figure}[h!]
	\centering
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=\linewidth]{figS5a.pdf}
		\caption{$\varphi_{\text{old}}$}
	\end{subfigure}%
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=\linewidth]{figS5b.pdf}
		\caption{$\varphi_{\text{new}}$}
	\end{subfigure}
	\caption{\textbf{Probability of adaptation for a large frequency of old-habitat patches ($f_{\text{old}}=0.9$).} The fecundity of the mutant in old-habitat patches is set to $\omega^\text{old}_m=1.45$. Note also the difference between the scales of the y-axes in the two panels.}
	\label{Sfig:vary_f}
\end{figure}

>>>>

The probability of adaptation $P_{\mathrm{adapt}}$ calculated with eq.~\ref{eq:source_sink} is compared to simulation results in Fig.~\ref{fig:source_sink}. In spite of our approximations, the fit to simulation results is still very good. 
\linelabel{R1-33}As $P_{\mathrm{adapt}}$ includes the probabilities of establishment $\varphi_{k}$, here again, the shapes of the curves as function of the emigration probability $m$ depend on the fecundity of the mutant in old-habitat patches, $\omega^\text{old}_m$ (Figs.~\ref{fig:source_sink_m-35}, \ref{fig:source_sink_m-45}). Likewise, the rankings of the different dispersal schemes are comparable to the ones observed for the establishment probability. \linelabel{R2-35}The small differences in the ranking of the dispersal schemes between panels~\ref{fig:vary_m_est_old-45} and \ref{fig:source_sink_m-45} can be explained by the mutational input. Larger stationary populations sizes of the wild type in new habitats, $\widehat{N}_w^{\text{new}}$ result in more mutations during the observed time frame. This population size is altered by the different dispersal schemes, most importantly it is largest for the wild type having a bias towards old habitats, i.e. the Old-Old and the Old-New dispersal schemes \florence{It does not make intuitive sense why $\widehat{N}_w^{\text{new}}$ is higher with WT biased towards old... maybe then figures with $\widehat{N}_w^{\text{old}}$ and $\widehat{N}_w^{\text{new}}$ would help}\florence{edit: I plotted $\widehat{N}_w^{\text{new}}$ and it is not largest for WT preferring old.. \url{https://github.com/flodebarre/PeteRescue/tree/master/code}}.

@@

We identified two types of effects of the different dispersal schemes. First, by changing population densities in both habitat types, the dispersal schemes alter the growth rate of the mutant in old-habitat patches $a_{\text{old}}$. This is the primary reason for the ranking of the dispersal schemes\footnote{\florence{really need a figure!}}. Secondly, the different dispersal schemes affect the number of mutations arising in either habitat type. This has a minor effect on the probability of evolutionary rescue for the explored parameter range but is relevant when studying the origin of the successful mutant lineage \chg{(see also Fig.~\ref{Sfig:natal_habitat})}. As the genetic background may vary across patches, the origin of a successful mutation will also affect which mutations will hitchhike with it. \linelabel{R1-54}Similarly, in the case of polygenic rescue or under recombination \citep[e.g.][]{uecker_2015}, the origin of a mutant is likely to affect its success.

@@

\florence{[add here figure of $\widehat{N}_w^k$ as function of $m$]}

@@

\florence{[add here multi-panel figure of $\tilde{N}_w^k$ as function of $m$]}

@@

The mean reproduction matrix $\mathcal{M}$ of a mutant gives the average number of offspring in a certain habitat, dependent on the habitat type in which the mutant resides (see also eq.~\eqref{eq:lambda} in the main text):
\begin{equation}\label{Seq:mean_repro}
\mathcal{M} = \bordermatrix{ ~ & \text{old patch} & \text{new patch} \cr
	\text{old patch} & \left(1 - m \mu_m^{\text{new}} \right) (1+a_{\text{old}}) & m \mu_m^{\text{new}} (1+a_{\text{new}}) \cr
	\text{new patch} & m \mu_m^{\text{old}} (1+a_{\text{old}}) & \left(1 - m \mu_m^{\text{old}}\right) (1+a_{\text{new}})
}\ ,
\end{equation}
where the rows denote the parent locations, and the columns the patch type of the offspring.

Our goal is to apply Theorem 5.6 from \citet{haccou_book} which states that for a slightly super-critical branching process, i.e. where the survival probability is slightly above zero, the establishment probability can be expressed in terms of the largest eigenvalue $\rho$ and the corresponding left- and right-eigenvectors of the mean reproduction matrix $\mathcal{M}$, denoted by $u$ and $v$, respectively. The eigenvectors should be normalized in the following way: $u_1+u_2 = 1$ and $\sum_{i=1}^2 u_i v_i = 1$. The establishment probabilities are then given by 
\begin{equation}\label{Seq:theory}
\varphi_i = \frac{2(\rho-1)}{B} v_i + O(\varepsilon), 
\end{equation}
with
\begin{equation} 
B = \sum_{i=1}^2 u_i \sum_{j=1}^2 v_j \mathcal{M}_{ij} + \rho(1-\rho) \sum_{j=1}^2 u_j v_j^2\, . 
\end{equation}

\subsection{Computing the largest eigenvalue}
We first approximate the largest eigenvalue of $\mathcal{M}$ denoted by $\rho$. It is given by (see \textit{Mathematica} notebook)
\begin{equation}
\begin{aligned}
\rho &= \frac{1}{2}\left(2 + a_{\text{old}} + a_{\text{new}} - m -  m_m^{\text{new}} a_{\text{old}} - m_m^{\text{old}} a_{\text{new}} \right.\\
&\quad \left. + \sqrt{4 (m -1) (1+a_{\text{old}})(1+a_{\text{new}}) + (2+a_{\text{old}} + a_{\text{new}} - m - m_m^{\text{new}} a_{\text{old}} - m_m^{\text{old}} a_{\text{new}}})^2 \right)
\end{aligned}
\end{equation}
%
In order to make analytical progress and to identify under which conditions the process is slightly super-critical, i.e. $\rho>1$, we rescale the parameters by a small parameter $\varepsilon$. We set $a_{\text{old}} = \varepsilon, a_{\text{new}} = \varepsilon \xi$ and $m = \varepsilon\mu$. Assuming that $\varepsilon$ is small enough, i.e. effectively a weak selection assumption in old-habitat patches, we can neglect higher orders of $\varepsilon$ and find
\begin{equation}\label{eq:eigenvalue}
\begin{aligned}
\rho &\approx 1+\frac{1}{2}\varepsilon\left(1+ \xi - \mu + \sqrt{(\xi -1 + \mu_m^{\text{new}})^2 + 2(1-\xi+\mu_m^{\text{new}})\mu_m^{\text{old}} + (\mu_m^{\text{old}})^2  }\right)\\
&= 1+\frac{1}{2}\varepsilon\left(1+ \xi - \mu + \sqrt{\frac{\gamma}{1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}}}  }\right),
\end{aligned}
\end{equation}
where $\gamma$ is the rescaled version of the constant $C$ in the main text (eq.~\eqref{eq:normalization}), i.e.
\begin{equation}
\gamma = (1-f_{\text{old}})(\xi-1+\mu)^2 +  e^{\pi_m} f_{\text{old}} (\xi-1-\mu)^2\, .
\end{equation}
%
For $\varepsilon\to 0$, we find that $\rho\to 1$ (eq.~\eqref{eq:eigenvalue}) which means that the branching is slightly super-critical if $\rho>1$ and real. A sufficient condition for this to be true is that the first term in the parentheses of \eqref{eq:eigenvalue} is positive:
\begin{equation}
1+\xi-\mu > 0 \quad \Leftrightarrow \quad a_{\text{old}} + a_{\text{new}} - m > 0
\end{equation}
%
In case that the branching process is not super-critical the establishment probability in eq.~\eqref{Seq:theory} becomes negative and as such is not a probability anymore. Hence, we can simply reject negative solutions of the establishment probability and by that implicitly justify that our approximation is valid.%\\

@@

\subsection{Computing the establishment probability}
Eq.~\eqref{Seq:theory} involves normalized eigenvectors. Their precise form is of not much insight. We therefore omit stating them explicitly but refer to the \textit{Mathematica} notebook. Solving eq.~\eqref{Seq:theory} to the first order of $\varepsilon$ and transforming back to the original variables, we obtain after reordering the equation presented in the main text (eq.~\eqref{eq:estab_approx})

\subsection{Disentangling the contributions to the establishment probability\label{sec:SI_contributions}}

\florence{I know you spent a lot on time on this, so I am not lightly making the suggestion but... I think this subsection could be removed. Intuitive explanations are given in the main text. The separation in three terms is not super intuitive, and their interpretation is more complicated that may seem, because $a_{\text{old}}$ is not constant. Plotting $a_{\text{old}}$ already helps a lot}
\chg{We now proceed to explain} the three regions of the establishment probability from Fig.~\ref{fig:vary_m_est}(a) in the main text. These were defined by: (i) an initial increase of the establishment probability at low dispersal rates $m$; (ii) a local maximum with a subsequent decrease of the establishment probability; (iii) an increase of the establishment probability for high dispersal rates.

\begin{figure}[t!]
	\centering
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=\linewidth]{fig2a.pdf}
		\caption{$\omega^\text{old}_m=1.35$ (large fecundity difference)}
	\end{subfigure}%
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=\linewidth]{fig2c.pdf}
		\caption{$\omega^\text{old}_m=1.45$ (small fecundity difference)}
	\end{subfigure}
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=\linewidth]{figS1c.pdf}
		\caption{$\omega^\text{old}_m=1.35$ (large fecundity difference)}
	\end{subfigure}%
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=\linewidth]{figS1d.pdf}
		\caption{$\omega^\text{old}_m=1.45$ (small fecundity difference)}
	\end{subfigure}
	\caption{\textbf{Contribution of the different terms in eq.~\eqref{eq:estab_approx} to the establishment probability $\varphi_{\text{old}}$.} Subfigures (a,b) are the same as Figs.~\ref{fig:vary_m_est}(a,c) in the main text. They show the establishment probability for a single mutant individual arising in an old-habitat patch for varying emigration probabilities $m$. In subfigures (c,d) we plot the terms from eq.~\eqref{eq:estab_approx} separately ($\pi_w=0.5,\pi_m=-0.5$). Term 1, the mutant growth rate in old habitats (solid), increases with increasing dispersal rates as a consequence of relaxed competition. Term 2, the environmental effect (dashed), captures the differences between the growth rates in the habitats. The larger the difference, the larger its contribution to the overall establishment probability. Term 3, the effect of dispersal (dotted), (largely) increases with increasing dispersal rates $m$. The sum of the three terms is plotted as a dash-dotted line.}
	\label{Sfig:contribution}
\end{figure}

For clarity, we re-plot Figs.~\ref{fig:vary_m_est}(a,c) in Fig.~\ref{Sfig:contribution}(a,b), respectively. We try to explain the ongoing processes for each of the regions through the approximations of the establishment probabilities in eq.~\eqref{eq:estab_approx}, see also Fig.~\ref{Sfig:contribution}(c,d). Note that these explanations are only valid for the establishment probability of a mutant initially in an old-habitat patch, $\varphi_{\text{old}}$. For the intuition behind the shape of $\varphi_{\text{new}}$ we refer to the corresponding section in the main text.

Region (i) is explained by the positive effect of dispersal. Mutants disperse from old- to new-habitat patches where they \chg{have a higher growth rate}. This effect is mediated through the third term of the establishment probability in eq.~\eqref{eq:estab_approx}. \chg{Note that the first two terms of the approximation cancel out for small emigration probabilities $m$.}
While the third term increases with increasing emigration rate $m$, the second term in eq.~\eqref{eq:estab_approx} decreases, cf. Fig.~\ref{Sfig:contribution}. 
In the formula this is mediated through \chg{the increase of the local growth rate in the old-habitat, $a_{\text{old}}$, i.e.\ it becomes less negative. Then both factors of the second term, $a_{\text{old}}$ and the difference $(a_{\text{old}}-a_{\text{new}})$, increase (which in turn decreases term two). The }\chg{intuitive} reason is that due to larger emigration probabilities $m$, more individuals leave old habitats before the reproductive event. This relaxes competition in these patches and therefore increases the local growth rate of mutants in old-habitat patches.   
%
%The intuition is that habitat differences become less important under stronger population mixing -- recall that the dispersal rate $m$ appears in the constant $C$ in the denominator. This alone would explain the decreasing curve of the second term. Additionally though, the local growth rate in old-habitat patches, $a_{\text{old}}$, is linked to the population demography which changes for varying $m$. With larger dispersal rates, more individuals leave the densely populated old habitats. This results in alleviated competition pressure for the remaining individuals, thus increasing the local growth rate $a_{\text{old}}$. This decreases the difference of growth rates $a_{\text{old}}-a_{\text{new}}$. Therefore, the local maximum can be explained by the decreasing environmental influence and the increasing dispersal effect, the second and the third term in eq.~\eqref{eq:estab_approx}, respectively. Hence, region (ii), beginning with the local maximum, is dominated by the decreasing effect of the environment on the establishment probability. 
Finally, in region (iii) dispersal is so large that the population homogenizes. This results in even less competitive pressure in old-habitat patches. Eventually, this yields a positive growth rate $a_{\text{old}}$ (first term in eq.~\eqref{eq:estab_approx}). Therefore, this region is driven by the local growth rate in old habitats.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Native habitat type %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage
\renewcommand{\theequation}{C\arabic{equation}}
\setcounter{equation}{0}  % reset counter

@@

\florence{I don't understand this section, which does not do much besides announcing a plot... Remove? Unless AmNat wants figures in appendix sections (in which case all figures need to be moved)?}

@@

\section{Probability of establishment for large frequencies of old-habitat patches}
\florence{The figure is not referred to in the main text. Is this because we can remove it...?}
\chg{We plot the probability of establishment for a large frequency of old-habitat patches, $f_{\text{old}}$. As visible in Fig.~\ref{Sfig:vary_f} below, for high frequencies of old-habitat patches the probability of adaptation becomes very small, if not zero, for large emigration probabilities $m$. These high frequency of old-habitat patches are the patch configurations which mutants that were present before the first patch degrades, i.e. standing genetic variation mutants, experience. Therefore, it is very unlikely that the mutant, which will eventually rescue the population, was already present before the first degradation event.} This supports the explanation that homogenizing the population (large $m$) reduces the impact of standing genetic variation on the probability of evolutionary rescue, see Fig.~\ref{fig:sgv}(b) in the main text where $\pi_w=\pi_m=0$ was plotted.

\florence{y axis label in panel b needs to be corrected}
\begin{figure}[h!]
	\centering
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=\linewidth]{figS5a.pdf}
		\caption{$\varphi_{\text{old}}$}
	\end{subfigure}%
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=\linewidth]{figS5b.pdf}
		\caption{$\varphi_{\text{new}}$}
	\end{subfigure}
	\caption{\textbf{Probability of adaptation for a large frequency of old-habitat patches ($f_{\text{old}}=0.9$).} The fecundity of the mutant in old-habitat patches is set to $\omega^\text{old}_m=1.45$. Note also the difference between the scales of the y-axes in the two panels.}
	\label{Sfig:vary_f}
\end{figure}
